---
title: "Finding Earth 2.0: Predicting Exoplanet Habitability"
execute:
  echo: false
---

Eccentric orbits matter because they often make a planet’s climate less stable over time, which can reduce the long-term chances that liquid water and a habitable environment would persist. In that sense, eccentricity is often treated as a warning sign for habitability. To that end, this data analysis project involves the use of a binary logistic regression model to explore whether fundamental stellar system properties can help predict whether exoplanet orbits are eccentric or not. Through this predictive model, I hoped to construct a reliable way to sort exoplanets by orbital eccentricity, allowing astronomers to focus on exoplanets with circular orbits that are more likely to be habitable.

Further details on the construction and training of the model can be found under the Methodology tab.

## Results

```{r}
#| message: false
#| warning: false
library(tidyverse)

raw_data <- read_csv(
  "PSCompPars_2026.07.29_10.50.24.csv",
  skip = 328,
  col_types = cols(sy_icmagerr2 = col_double()),
  show_col_types = FALSE
)

if ("janitor" %in% rownames(installed.packages())) {
  library(janitor)
  raw_data <- clean_names(raw_data)
}

df_clean <- raw_data %>%
  mutate(
    is_eccentric = as.integer(pl_orbeccen >= 0.1),
    st_mass_logg = st_mass * st_logg
  ) %>%
  select(
    is_eccentric,
    st_met,
    st_mass,
    st_teff,
    st_logg,
    st_mass_logg,
    sy_pnum
  ) %>%
  drop_na()

set.seed(123)
train_idx <- sample(seq_len(nrow(df_clean)), size = floor(0.8 * nrow(df_clean)))
train_data <- df_clean[train_idx, ]
test_data <- df_clean[-train_idx, ]

logit_model <- glm(
  is_eccentric ~ st_met + st_mass + st_teff + st_logg + st_mass_logg + sy_pnum,
  data = train_data,
  family = binomial(link = "logit")
)

test_plot_data <- test_data %>%
  mutate(
    predicted_prob = predict(logit_model, newdata = test_data, type = "response"),
    predicted_class = as.integer(predicted_prob >= 0.26),
    actual_class = factor(is_eccentric, levels = c(0, 1), labels = c("Circular", "Eccentric")),
    prediction_match = predicted_class == is_eccentric
  )
```

The figure below summarizes the estimated effects of the variables in the model. Values above 1 indicate higher odds of an eccentric orbit, while values below 1 indicate lower odds, holding the other variables fixed. The estimated odds ratio for metallicity is about 2.39, with a 95% confidence interval from about 1.55 to 3.54. The estimated odds ratio for surface gravity is about 0.35, with a 95% confidence interval from about 0.29 to 0.44. I also included an interaction term between stellar mass and surface gravity to test whether their combined effect differs from their separate effects.

```{r}
#| echo: false
#| warning: false
odds_results <- as.data.frame(exp(cbind(OR = coef(logit_model), confint(logit_model)))) %>%
  rownames_to_column("term") %>%
  filter(term != "(Intercept)") %>%
  rename(
    odds_ratio = OR,
    lower = `2.5 %`,
    upper = `97.5 %`
  ) %>%
  mutate(
    term = recode(
      term,
      st_met = "Stellar metallicity",
      st_mass = "Stellar mass",
      st_teff = "Stellar temperature",
      st_logg = "Stellar surface gravity",
      st_mass_logg = "Stellar mass × surface gravity",
      sy_pnum = "Planets in system"
    ),
    term = factor(
      term,
      levels = rev(c(
        "Stellar metallicity",
        "Stellar mass",
        "Stellar temperature",
        "Stellar surface gravity",
        "Stellar mass × surface gravity",
        "Planets in system"
      ))
    )
  )

print(
  ggplot(odds_results, aes(x = term, y = odds_ratio)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "#7f8c8d", linewidth = 1, alpha = 0.8) +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = "#2563eb", linewidth = 1.1) +
    geom_point(size = 3.6, color = "#2563eb", fill = "#dbeafe", shape = 21, stroke = 1.1) +
    coord_flip() +
    scale_y_log10(limits = c(0.2, 4.5), breaks = c(0.25, 0.5, 1, 2, 4)) +
    labs(
      title = "Odds ratios for the model covariates",
      x = NULL,
      y = "Odds ratio (log scale)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      panel.grid.major.y = element_line(color = "#e5e7eb", linewidth = 0.7),
      panel.grid.major.x = element_line(color = "#f3f4f6", linewidth = 0.6),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 16, color = "#111827"),
      axis.text = element_text(color = "#374151", size = 12),
      axis.title = element_text(color = "#374151", size = 13),
      plot.background = element_rect(fill = "#f8fafc", color = NA),
      panel.background = element_rect(fill = "#f8fafc", color = NA),
      axis.line = element_blank()
    )
)
```

Metallicity and surface gravity were the clearest predictors in this model. The interaction term between stellar mass and surface gravity was small and not statistically significant.

```{r}
#| echo: false
#| warning: false
# Compute precision and recall across many thresholds
thresholds <- sort(unique(c(0, 1, seq(0, 1, by = 0.01), test_plot_data$predicted_prob)))

pr_df <- map_dfr(thresholds, function(threshold) {
  predicted_positive <- test_plot_data$predicted_prob >= threshold
  tp <- sum(predicted_positive & test_plot_data$is_eccentric == 1)
  fp <- sum(predicted_positive & test_plot_data$is_eccentric == 0)
  fn <- sum(!predicted_positive & test_plot_data$is_eccentric == 1)

  tibble(
    threshold = threshold,
    precision = ifelse(tp + fp == 0, NA_real_, tp / (tp + fp)),
    recall = ifelse(tp + fn == 0, NA_real_, tp / (tp + fn))
  )
}) %>%
  filter(!is.na(precision), !is.na(recall)) %>%
  arrange(recall, precision)

print(
  ggplot(pr_df, aes(x = recall, y = precision)) +
    geom_line(color = "#2563eb", linewidth = 1.4) +
    geom_area(fill = "#93c5fd", alpha = 0.25) +
    geom_hline(yintercept = mean(test_plot_data$is_eccentric == 1), linetype = "dashed", color = "#7f8c8d", linewidth = 1) +
    scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
    labs(
      title = "Precision-recall curve",
      x = "Recall",
      y = "Precision"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      panel.grid.major = element_line(color = "#e5e7eb", linewidth = 0.7),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 16, color = "#111827"),
      axis.text = element_text(color = "#374151", size = 12),
      axis.title = element_text(color = "#374151", size = 13),
      plot.background = element_rect(fill = "#f8fafc", color = NA),
      panel.background = element_rect(fill = "#f8fafc", color = NA),
      axis.line = element_blank()
    )
)
```

This precision-recall curve summarizes the trade-off between precision and recall across a range of thresholds. The curve indicates that the model has limited ability to identify eccentric cases while avoiding false positives.

## Conclusion

This project shows that some broad star-system properties can offer clues about whether a planet is more likely to have an eccentric orbit. In particular, stellar metallicity and surface gravity stand out as the strongest predictors in the model. The interaction term between stellar mass and surface gravity did not add much to the model. The results suggest that basic star-system measurements provide useful information, but they do not fully determine eccentricity.
