---
title: "Results"
execute:
  echo: false
---

## Model results

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(plotly)

raw_data <- read_csv(
  "PSCompPars_2026.07.29_10.50.24.csv",
  skip = 328,
  col_types = cols(sy_icmagerr2 = col_double()),
  show_col_types = FALSE
)

if ("janitor" %in% rownames(installed.packages())) {
  library(janitor)
  raw_data <- clean_names(raw_data)
}

df_clean <- raw_data %>%
  mutate(
    is_eccentric = as.integer(pl_orbeccen >= 0.1)
  ) %>%
  select(
    is_eccentric,
    st_met,
    st_mass,
    st_teff,
    st_logg,
    sy_pnum
  ) %>%
  drop_na()

set.seed(123)
train_idx <- sample(seq_len(nrow(df_clean)), size = floor(0.8 * nrow(df_clean)))
train_data <- df_clean[train_idx, ]
test_data <- df_clean[-train_idx, ]

logit_model <- glm(
  is_eccentric ~ st_met + st_mass + st_teff + st_logg + sy_pnum,
  data = train_data,
  family = binomial(link = "logit")
)

test_plot_data <- test_data %>%
  mutate(
    predicted_prob = predict(logit_model, newdata = test_data, type = "response"),
    predicted_class = as.integer(predicted_prob >= 0.26),
    actual_class = factor(is_eccentric, levels = c(0, 1), labels = c("Circular", "Eccentric")),
    prediction_match = predicted_class == is_eccentric
  )
```

### Effect estimates
- Metal-rich stars were more likely to host planets with eccentric orbits in this dataset. In practical terms, moving up in metallicity was linked to a noticeably higher chance of eccentricity.
- Stars with higher surface gravity were less likely to be linked with eccentric orbits, suggesting those systems may be dynamically calmer.
- Because these two star properties are measurable early, they can help prioritize which systems deserve closer follow-up.

The figure below shows which inputs push predictions toward eccentric or circular outcomes. Points above 1 mean a variable tends to raise the chance of eccentricity; points below 1 mean it tends to lower it. The horizontal bars show uncertainty, so wider bars mean less confidence in the exact effect size.

```{r}
#| echo: false
#| warning: false
odds_results <- exp(cbind(OR = coef(logit_model), confint(logit_model))) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  filter(term != "(Intercept)") %>%
  rename(
    odds_ratio = OR,
    lower = `2.5 %`,
    upper = `97.5 %`
  ) %>%
  mutate(
    term = recode(
      term,
      st_met = "Stellar metallicity",
      st_mass = "Stellar mass",
      st_teff = "Stellar temperature",
      st_logg = "Stellar surface gravity",
      sy_pnum = "Planets in system"
    ),
    term = factor(
      term,
      levels = c(
        "Planets in system",
        "Stellar surface gravity",
        "Stellar temperature",
        "Stellar mass",
        "Stellar metallicity"
      )
    )
  )

print(
  ggplot(odds_results, aes(x = term, y = odds_ratio)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "#7f8c8d", linewidth = 1, alpha = 0.8) +
    geom_linerange(aes(ymin = lower, ymax = upper), color = "#5ec9ff", linewidth = 1.1) +
    geom_point(size = 3.6, color = "#5ec9ff", fill = "#10162b", shape = 21, stroke = 1.1) +
    coord_flip() +
    scale_y_log10(limits = c(0.2, 4.5), breaks = c(0.25, 0.5, 1, 2, 4)) +
    labs(
      title = "Odds ratios for the model covariates",
      x = NULL,
      y = "Odds ratio (log scale)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      text = element_text(family = "Segoe UI", color = "#f4f7ff"),
      panel.grid.major.y = element_line(color = "#2e3a58", linewidth = 0.7),
      panel.grid.major.x = element_line(color = "#25324a", linewidth = 0.6),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 16, color = "#f4f7ff"),
      axis.text = element_text(color = "#dbeafe", size = 12),
      axis.title = element_text(color = "#f4f7ff", size = 13),
      plot.background = element_rect(fill = "#0b1324", color = NA),
      panel.background = element_rect(fill = "#0b1324", color = NA),
      axis.line = element_blank()
    )
)
```

Metallicity and surface gravity were the clearest predictors in this model.

### Prediction quality

The next figures focus on how the model separates the two classes and how well it performs on the held-out test set. In this setting, the precision-recall curve is especially informative because it highlights the balance between recovering genuinely eccentric systems and avoiding a large number of false alarms.

```{r}
#| echo: false
#| warning: false
thresholds <- sort(unique(c(0, 1, seq(0, 1, by = 0.01), test_plot_data$predicted_prob)))

pr_df <- map_dfr(thresholds, function(threshold) {
  predicted_positive <- test_plot_data$predicted_prob >= threshold
  tp <- sum(predicted_positive & test_plot_data$is_eccentric == 1)
  fp <- sum(predicted_positive & test_plot_data$is_eccentric == 0)
  fn <- sum(!predicted_positive & test_plot_data$is_eccentric == 1)

  tibble(
    threshold = threshold,
    precision = ifelse(tp + fp == 0, NA_real_, tp / (tp + fp)),
    recall = ifelse(tp + fn == 0, NA_real_, tp / (tp + fn))
  )
}) %>%
  filter(!is.na(precision), !is.na(recall)) %>%
  arrange(recall, precision)

pr_auc <- sum(
  diff(pr_df$recall) * (head(pr_df$precision, -1) + tail(pr_df$precision, -1)) / 2,
  na.rm = TRUE
)

baseline_precision <- mean(test_plot_data$is_eccentric == 1)

pr_plot <- plot_ly(
  pr_df,
  x = ~recall,
  y = ~precision,
  type = "scatter",
  mode = "lines",
  line = list(color = "#5ec9ff", width = 2.5),
  fill = "tozeroy",
  fillcolor = "rgba(94, 201, 255, 0.18)",
  customdata = ~threshold,
  hovertemplate = paste0(
    "Recall: %{x:.3f}<br>",
    "Precision: %{y:.3f}<br>",
    "Threshold: %{customdata:.2f}<extra></extra>"
  ),
  name = "PR curve"
) %>%
  add_segments(
    x = 0,
    xend = 1,
    y = baseline_precision,
    yend = baseline_precision,
    inherit = FALSE,
    line = list(color = "#d6b4ff", width = 1.2, dash = "dash"),
    hovertemplate = paste0(
      "Random-baseline precision: ",
      scales::number(baseline_precision, accuracy = 0.001),
      "<extra></extra>"
    ),
    name = "Baseline"
  ) %>%
  layout(
    title = list(text = "Precision-recall curve", font = list(color = "#f4f7ff", size = 18)),
    xaxis = list(
      title = "Recall",
      range = c(0, 1),
      tickvals = seq(0, 1, by = 0.25),
      color = "#dbeafe",
      gridcolor = "#2e3a58",
      zeroline = FALSE
    ),
    yaxis = list(
      title = "Precision",
      range = c(0, 1),
      tickvals = seq(0, 1, by = 0.25),
      color = "#dbeafe",
      gridcolor = "#2e3a58",
      zeroline = FALSE
    ),
    paper_bgcolor = "#0b1324",
    plot_bgcolor = "#0b1324",
    legend = list(font = list(color = "#f4f7ff")),
    hoverlabel = list(bgcolor = "#10162b", font = list(color = "#f4f7ff"))
  )

pr_plot
```

This precision-recall curve shows the trade-off between catching more eccentric planets and making more false alarms.

- AUPRC (area under the precision-recall curve) is `r round(pr_auc, 3)`. It summarizes the whole curve into one score: higher means the model keeps true detections high while limiting false positives across many thresholds.
- Why AUPRC matters here: eccentric planets are the minority class, so AUPRC is more informative than accuracy alone for judging whether the model is truly useful for discovery.
- The dashed baseline (`r scales::percent(mean(test_plot_data$is_eccentric == 1), accuracy = 0.1)`) is what you would expect from random guessing at the class prevalence. Performance above that line means the model is adding real signal.
- At threshold 0.26, precision is `r scales::percent(sum(test_plot_data$predicted_class == 1 & test_plot_data$is_eccentric == 1) / sum(test_plot_data$predicted_class == 1), accuracy = 0.1)` and recall is `r scales::percent(sum(test_plot_data$predicted_class == 1 & test_plot_data$is_eccentric == 1) / sum(test_plot_data$is_eccentric == 1), accuracy = 0.1)`: the model finds some eccentric systems, but misses many and still raises false alarms.

```{r}
#| echo: false
#| warning: false
metric_df <- tibble(
  metric = c("Accuracy", "Precision", "Sensitivity", "Specificity"),
  value = c(
    mean(test_plot_data$prediction_match),
    sum(test_plot_data$predicted_class == 1 & test_plot_data$is_eccentric == 1) / sum(test_plot_data$predicted_class == 1),
    sum(test_plot_data$predicted_class == 1 & test_plot_data$is_eccentric == 1) / sum(test_plot_data$is_eccentric == 1),
    sum(test_plot_data$predicted_class == 0 & test_plot_data$is_eccentric == 0) / sum(test_plot_data$is_eccentric == 0)
  )
)

print(
  ggplot(metric_df, aes(x = reorder(metric, -value), y = value)) +
    geom_col(fill = "#6d7cff", alpha = 0.95) +
    geom_text(aes(label = scales::percent(value, accuracy = 0.1)), vjust = -0.4, size = 3.5, color = "#f4f7ff") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = "Test-set classification metrics",
      x = NULL,
      y = "Value"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      text = element_text(family = "Segoe UI", color = "#f4f7ff"),
      panel.grid.major.y = element_line(color = "#2e3a58", linewidth = 0.7),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 16, color = "#f4f7ff"),
      axis.text = element_text(color = "#dbeafe", size = 12),
      axis.title = element_text(color = "#f4f7ff", size = 13),
      plot.background = element_rect(fill = "#0b1324", color = NA),
      panel.background = element_rect(fill = "#0b1324", color = NA)
    )
)
```

This bar chart gives four different views of model quality, each answering a different practical question.

- Accuracy asks: how often is the model correct overall? This can look better than the model really is when one class is much more common.
- Precision asks: when the model flags an eccentric orbit, how often is that flag correct?
- Sensitivity (recall) asks: out of truly eccentric systems, how many did the model successfully catch?
- Specificity asks: out of truly circular systems, how many did the model correctly leave unflagged?

```{r}
#| echo: false
#| warning: false
class_balance <- df_clean %>%
  mutate(
    class_label = factor(is_eccentric, levels = c(0, 1), labels = c("Circular", "Eccentric"))
  ) %>%
  count(class_label) %>%
  mutate(proportion = n / sum(n))

print(
  ggplot(class_balance, aes(x = class_label, y = proportion, fill = class_label)) +
    geom_col(alpha = 0.95) +
    geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)), vjust = -0.4, size = 3.5, color = "#f4f7ff") +
    scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = c("#6d7cff", "#5ec9ff")) +
    labs(
      title = "Class balance in the cleaned dataset",
      x = NULL,
      y = "Proportion of observations"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      text = element_text(family = "Segoe UI", color = "#f4f7ff"),
      panel.grid.major.y = element_line(color = "#2e3a58", linewidth = 0.7),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 16, color = "#f4f7ff"),
      axis.text = element_text(color = "#dbeafe", size = 12),
      axis.title = element_text(color = "#f4f7ff", size = 13),
      plot.background = element_rect(fill = "#0b1324", color = NA),
      panel.background = element_rect(fill = "#0b1324", color = NA),
      legend.position = "none"
    )
)
```

In the cleaned dataset, circular orbits are much more common (`r scales::percent(mean(df_clean$is_eccentric == 0), accuracy = 0.1)`).

- The model therefore sees many more circular examples during training.
- As a result, performance can look strong even when many of the rarer eccentric systems are still missed.

```{r}
#| echo: false
#| warning: false
prob_density_plot_data <- test_plot_data %>%
  mutate(
    actual_class = factor(is_eccentric, levels = c(0, 1), labels = c("Circular", "Eccentric"))
  )

density_df <- prob_density_plot_data %>%
  group_by(actual_class) %>%
  group_modify(~ {
    d <- density(.x$predicted_prob, from = 0, to = 1, n = 512)
    tibble(predicted_prob = d$x, density = d$y)
  }) %>%
  ungroup()

density_colors <- c("Circular" = "#6d7cff", "Eccentric" = "#5ec9ff")

density_plot <- plot_ly()
for (cls in names(density_colors)) {
  class_df <- density_df %>% filter(actual_class == cls)
  density_plot <- density_plot %>%
    add_trace(
      data = class_df,
      x = ~predicted_prob,
      y = ~density,
      type = "scatter",
      mode = "lines",
      fill = "tozeroy",
      name = cls,
      line = list(color = density_colors[[cls]], width = 2),
      fillcolor = if (cls == "Circular") "rgba(109, 124, 255, 0.35)" else "rgba(94, 201, 255, 0.35)",
      hovertemplate = paste0(
        "Class: ", cls, "<br>",
        "Predicted probability: %{x:.3f}<br>",
        "Density: %{y:.3f}<extra></extra>"
      )
    )
}

density_plot <- density_plot %>%
  layout(
    title = list(text = "Distribution of predicted probabilities", font = list(color = "#f4f7ff", size = 18)),
    xaxis = list(
      title = "Predicted probability of eccentricity",
      range = c(0, 1),
      tickvals = seq(0, 1, by = 0.25),
      color = "#dbeafe",
      gridcolor = "#2e3a58",
      zeroline = FALSE
    ),
    yaxis = list(
      title = "Density",
      color = "#dbeafe",
      gridcolor = "#2e3a58",
      zeroline = FALSE
    ),
    paper_bgcolor = "#0b1324",
    plot_bgcolor = "#0b1324",
    legend = list(title = list(text = "Observed class"), font = list(color = "#f4f7ff")),
    hoverlabel = list(bgcolor = "#10162b", font = list(color = "#f4f7ff"))
  )

density_plot
```

This density plot compares the predicted probabilities for circular and eccentric cases. A better separation between the two groups would show as two more distinct peaks.

```{r}
#| echo: false
#| warning: false
confusion_data <- test_plot_data %>%
  mutate(
    actual_class = factor(is_eccentric, levels = c(0, 1), labels = c("Actual: circular", "Actual: eccentric")),
    predicted_class_label = factor(predicted_class, levels = c(0, 1), labels = c("Predicted: circular", "Predicted: eccentric"))
  ) %>%
  count(actual_class, predicted_class_label) %>%
  complete(actual_class, predicted_class_label, fill = list(n = 0))

print(
  ggplot(confusion_data, aes(x = predicted_class_label, y = actual_class, fill = n)) +
    geom_tile(color = "#0b1324", linewidth = 1) +
    geom_text(aes(label = n), color = "#f4f7ff", size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#10162b", high = "#5ec9ff") +
    labs(
      title = "Confusion matrix for the test set",
      x = NULL,
      y = NULL,
      fill = "Count"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      text = element_text(family = "Segoe UI", color = "#f4f7ff"),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", size = 16, color = "#f4f7ff"),
      axis.text = element_text(color = "#dbeafe", size = 12),
      axis.title = element_text(color = "#f4f7ff", size = 13),
      plot.background = element_rect(fill = "#0b1324", color = NA),
      panel.background = element_rect(fill = "#0b1324", color = NA),
      legend.text = element_text(color = "#f4f7ff"),
      legend.title = element_text(color = "#f4f7ff")
    )
)
```

This confusion matrix is a direct count of right and wrong predictions in each class.

- Most circular systems are identified correctly, but many eccentric systems are still missed.
- For a screening tool, missed eccentric systems reduce discovery value, while false alarms increase costly follow-up work.

## Conclusion

### Key takeaways
- Stellar metallicity and surface gravity stand out as the strongest predictors in the model.
  - A plausible explanation is that metallicity reflects the chemical enrichment of the protoplanetary disk, which may influence how planets form and how dynamically complex their orbits become.
  - Surface gravity may proxy for stellar structure and evolutionary context, which can shape the broader architecture of the system in ways that affect orbital eccentricity.
- The results suggest that basic star-system measurements provide useful information, but they do not fully determine eccentricity.
  - This is unsurprising, because orbital eccentricity is also governed by dynamical processes such as planet-planet scattering, secular perturbations, and migration history.

### Limitations of the dataset and study
- The analysis relies on a relatively small and observationally derived dataset, so the conclusions should be viewed as suggestive rather than definitive.
- The binary outcome is a simplification of a continuous physical phenomenon; some systems near the threshold may be classified somewhat arbitrarily.
- The model uses only a small set of host-star and system variables, and it does not include other potentially important drivers such as orbital architecture, age, migration history, or planet size.
- Selection effects in exoplanet surveys may bias the sample toward systems that are easier to detect, which can influence the apparent relationships between stellar properties and orbital eccentricity.